Comparing the Direction and Degree of Selection on Mutations Between SARS-CoV-2 Phylogenetic Clades

### 2022-10-21 notebooks/expected_and_actual_counts_ratio.Rmd
Add commands to compute p-values using Fisher's Exact Test.

#install.packages("ggplot2")
#install.packages("stringr")
library(ggplot2)
library(stringr)
```

@@

```{r read_data}
SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE)
```

@@

#Add column with the mutation type (clade founder.mutation).
filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1),".",str_sub(filtered_SARS_mutation$nt_mutation,-1,-1),sep="")

#For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
```

@@

```{r contingency}

>>>>

#install.packages("ggplot2")
#install.packages("stringr")
#install.packages("gtools")

library(ggplot2)
library(stringr)
library(gtools)
```

Discuss purpose of notebook and steps to achieve that goal. Workflow to produce a volcano plot between any two clades.
List steps to do analysis (e.g., how to structure the data).
Do some sanity checks.

@@

```{r read_data}
SARS_mutation <- read.csv("../results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv", header=TRUE) # Read data from 'kfeldmann' computational notebook on server.
SARS_mutation <- read.csv("https://media.githubusercontent.com/media/jbloomlab/SARS2-mut-rates/main/results/expected_vs_actual_mut_counts/expected_vs_actual_mut_counts.csv") # Read data from github.com.
```

@@

#Add column with the mutation type (clade founder.mutation).
filtered_SARS_mutation$mutation_type <- paste(str_sub(filtered_SARS_mutation$nt_mutation,1,1), ".", str_sub(filtered_SARS_mutation$nt_mutation,-1,-1), sep="")

#For visualization purposes, reduce the number of bins by combining clades of the same Nextstrain variant.
#filtered_SARS_mutation$variant <- filtered_SARS_mutation$clade
#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="20I"] <- "Alpha"
#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21C"] <- "Epsilon"
#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21I" | filtered_SARS_mutation$variant=="21J"] <- "Delta"
#filtered_SARS_mutation$variant[filtered_SARS_mutation$variant=="21K" | filtered_SARS_mutation$variant=="21M" | filtered_SARS_mutation$variant=="22A" | filtered_SARS_mutation$variant=="22B"] <- "Omicron"
```

@@

```{r contingency}
# CREATE SUBSET OF DATA TO WORK WITH: ONLY CLADES 20A AND 20B FOR ALL COUNTRIES
temp <- filtered_SARS_mutation
filtered_SARS_mutation <- temp[which(temp$clade=="20A" | temp$clade=="20B"),]

# Filter mutation data for Fisher's Exact test, and combine nucleotide sites, mutation types and clades to make an iterative reference.
fisher_data <- data.frame(nt_mutation_clade=paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, filtered_SARS_mutation$clade, sep="_"), observed=filtered_SARS_mutation$actual_count, expected=filtered_SARS_mutation$expected_count)

# Identify unique clades in the dataset.
clades <- unique(filtered_SARS_mutation$clade)

# Create a two-column matrix with all of the possible clade-clade comparisons (each row contains one comparison).
clade_comparisons <- combinations(n=length(clades), r=2, v=clades, repeats.allowed=FALSE)

# Combine nucleotide sites and mutation types, and filter list to only include unique nucleotide mutations.
nt_mutation_list <- unique(paste(filtered_SARS_mutation$nt_site, filtered_SARS_mutation$mutation_type, sep="_"))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the nucleotide mutation list by the number of clade comparisons.
rep_nt_mutation <- data.frame(nt_mutation=rep(nt_mutation_list, each=nrow(clade_comparisons)))

# In preparation for combining clade comparisons with unique nucleotide mutations, repeat the clade comparison matrix by the number of unique nucleotide mutations.
rep_clades <- do.call("rbind",replicate(length(nt_mutation_list), clade_comparisons, simplify=FALSE))

# Combine the nucleotide mutations to each column in the clade comparison matrix to create a reference for the fisher data.
comparisons <- data.frame(clade1=paste(rep_nt_mutation$nt_mutation, rep_clades[,1], sep="_"), clade2=paste(rep_nt_mutation$nt_mutation, rep_clades[,2], sep="_"))

# Drop comparisons where the nucleotide site, mutation type, clade reference does not exist in the observed-expected data frame.
filtered_comparisons <- comparisons[comparisons$clade1 %in% fisher_data$nt_mutation_clade & comparisons$clade2 %in% fisher_data$nt_mutation_clade,]

# Save list of p-values for every comparison. NOTE: Expected counts are not integers and are therefore rounded for the Fisher's Exact test.
p_values <- data.frame(p_value=apply(filtered_comparisons, 1, FUN=function(c) {fisher.test(fisher_data[c(which(fisher_data$nt_mutation_clade==c[1]), which(fisher_data$nt_mutation_clade==c[2])), c("observed","expected")])$p.value}))

save(p_values,file="fisher_pvalues.Rda")
load("fisher_pvalues.Rda")

# Combine p-value list with the reference list of clade comparisons.

# run fishers exact an save p-value to dataframe (clade comparison, site, mutation, p-value, fdr p-value, bonferroni p-value)

# test run volcano plot for each clade comparison

# sanity check analyses (distribution of synonymous p-values)
```
